WASP-4 is Accelerating Towards the Sun

### 2020-02-04 paper/ms.tex
Update ms.tex

\title{WASP-4 is Accelerating Towards the Earth}
% ", Explaining its Decreasing Orbital Period"
% ": A Cautionary Tale in the Search for Tidal Orbital Decay"
% "An Explanation for WASP-4's Decreasing Orbital Period".
% "WASP-4's Decreasing Orbital Period: Solved."

@@

\begin{abstract}
  Space and ground-based transit measurements of the hot Jupiter
  WASP-4b have recently shown that its orbital period appears to be
  decreasing.  Proposed explanations for the period change include
  tidal orbital decay, orbital precession, and light-travel time
  effects.  We present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.  The data show that the system is
  accelerating towards the Earth at $\dot{\gamma} =
  -0.0422^{+0.0028}_{-0.0027}\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
  The implied period decrease explains most or all of WASP-4's
  changing orbital period as a light-travel time effect ($-5.94 \pm
  0.39~{\rm msec}\,{\rm yr}^{-1}$ implied, compared to $-8.64 \pm
  1.26~{\rm msec}\,{\rm yr}^{-1}$ observed).  Combining the radial
  velocities with new upper limits from speckle imaging, we find that
  the system's acceleration is likely caused by a 10-200$\,M_{\rm
  Jup}$ companion with semi-major axis between 10-100$\,$AU.  The
  statistics from \citet{knutson_friends_2014} imply that about 1 in 6
  hot Jupiters are expected to show period decreases comparable to
  WASP-4, due to acceleration by outer companions.  These period
  changes will be increasingly measureable in coming years, because
  the precision of the period derivative measurement scales with the
  observing baseline to the power of 5/2.  Continued radial velocity
  monitoring of hot Jupiters is therefore essential to distinguish
  tidal orbital decay from line-of-sight accelerations.
\end{abstract}

@@

The orbits of most hot Jupiters are unstable to tidal decay
\citep{counselman_outcomes_1973,hut_stability_1980,rasio_tidal_1996,levrard_falling_2009,matsumura_tidal_2010}.
The relevant issue is whether the timescale for tidal orbital decay is
shorter or longer than the timescale for main-sequence stellar
evolution.  This question depends on the uncertain rate at which
friction inside the star can damp the energy of tidal oscillations and
thereby shrink the orbit (as reviewed by \citealt{Mazeh2008} and
\citealt{ogilvie_tidal_2014}).

Indirect studies of age and angular momentum indicators including the
hot Jupiter semi-major axis distribution, host star spin rates, and
galactic velocity dispersions have led to estimates for the inspiral
timescale that vary from much less to much greater than the
main-sequence evolution time  ({\it e.g.},
\citealt{jackson_observational_2009}, \citealt{teitler_why_2014},
\citealt{penev_empirical_2018}, \citealt{cameron_hierarchical_2018},
\citealt{hamer_schlaufman_2019}).  Direct measurements of tidal
orbital decay through transit and occultation timing could provide an
empirical resolution.  For instance, combined transit timing and
radial velocity measurements for WASP-12b have shown that its secular
period decrease of $\approx$30 milliseconds per year is probably due
to tidal orbital decay
\citep{maciejewski_departure_2016,patra_2017,yee_orbit_2020}.

This study highlights a point that, though obvious, has perhaps not
yet received due attention.  The point is that observational programs
aimed at identifying orbital decay in hot Jupiters through transit
timing will be crippled without concurrent long-term radial velocity
monitoring.  The reason is that line-of-sight accelerations due to
outer companions \citep[{\it e.g.},][]{agol_detecting_2005} and tidal
orbital decay both initially manifest identically in transit times, as
a non-zero period derivative.  Massive outer companions to hot
Jupiters are the norm; \citet{bryan_statistics_2016} calculated an
occurrence rate of $60.9^{+5.2}_{-5.6}\%$ for outer companions to hot
Jupiters with masses from 1-20$\,$$M_{\rm Jup}$ and semi-major axes
from 5-100$\,$AU.  Therefore it could very well be that more hot
Jupiters will show shrinking orbital periods due to long term
accelerations than due to tidal orbital decay.

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that appears to be decreasing by about 10 milliseconds
per year.  We discovered the timing variations by combining data from
TESS \citep{ricker_transiting_2015} with a decade of ground-based
observations \citep[][hereafter
\citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.  Thereafter,
\citet{southworth_transit_2019} reported 22 new transit times for the
system, and found an updated decay rate of $\dot{P} = -9.2 \pm 1.1$
milliseconds per year. The \citeauthor{southworth_transit_2019} decay
rate was $\approx$3$\sigma$ less rapid than that found by
\citetalias{bouma_wasp4b_2019}, but the conclusions of the studies
were otherwise similar.  A separate study by \citet{baluev_2019}
reported additional archival transit light curves of WASP-4b from
TRAPPIST and select other observers.  \citet{baluev_2019} pointed out
that when using lower-precision subsets of the available transit data,
the case for a decreasing period worsened.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the five available
HIRES radial velocities suggested a weak ($\approx$$2\sigma$) linear
trend \citep{knutson_friends_2014}.  Our new measurements reveal a
line-of-sight acceleration of $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m\,s^{-1}\,day^{-1}}$.  This
translates to an expected period decrease from the light-travel time
effect of -5.9 milliseconds per year---about commensurate with what is
observed from transits.  While the apparent period change caused by a
line-of-sight acceleration has been referred to as the ``R{\o}mer
effect'' \citep{yee_orbit_2020}, we prefer to avoid this term, which
typically signifies arrival time delay due to observatory motion. We
are discussing the Doppler effect seen by a stationary observer for a
source with constant line-of-sight acceleration.

In the following, Section~\ref{sec:observations} collects the
available transit data and presents the new radial velocity and
speckle imaging observations.  Section~\ref{sec:analysis} analyzes the
data, and finds that they yield a picture in which the WASP-4 system
is accelerating towards our line-of-sight, likely due to the pull of a
brown or M-dwarf companion.  Section~\ref{sec:discussion} places this
result in the broader context of orbital decay searches, and points
out that line-of-sight accelerations, {\it i.e.}, ``false positive
orbital decay signals'', are relatively common in the hot Jupiter
population.  Section~\ref{sec:conclusions} offers concluding remarks.

@@

Table~1 lists the transit times we collected for our analysis.  We
include data from the peer-reviewed literature for which {\it (i)} the
analysis was based on observations of a single transit, {\it (ii)} the
midpoint was fitted as a free parameter, and {\it (iii)} the time
system specified both the leap second correction (TDB or UTC) and also
whether any barycentric or heliocentric corrections had been
performed.

% FIXME : verify with Michael Gillon that this is true. (P. Evans
% done)
The majority of times are identical to those we collected in
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m NTT and Danish 1.54m telescopes at La Silla,
and the SAAO 1.0m telescope.
% Five of the twenty-two occurred at simultaneous epochs as transits
% previously reported by \citet{hoyer_tramos_2013}.

Additional timing measurements were also recently made available by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations.
We included twelve of their ``high quality'' transit times from
TRAPPIST (six transits), El~Sauce (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that correct barycentric and leap-second
corrections had been performed (M.~Gillon, P.~Evans, priv{.} comm{.}).
We omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had or had not been performed.

The four available occultations tabulated by
\citetalias{bouma_wasp4b_2019} have neglegible statistical value due
to their large uncertainties, and we forgo their use in this analysis.

@@

After identifying the period decrease in
\citetalias{bouma_wasp4b_2019}, we acquired four additional radial
velocity measurements with the Keck High Resolution Echelle
Spectrometer (HIRES; \citealt{vogt_hires_1994}).  Our observations
were acquired using the standard setup and reduction techniques of the
California Planet Survey \citep{howard_cps_2010}.  Previously, the
HIRES data-points spanned 2010 to 2013 \citep{knutson_friends_2014}.
Our new measurements triple the HIRES observing baseline to nine
years.

The complete set of radial velocity observations is given in Table~2.
Along with the 2010-2019 HIRES observations, there are also
measurements from CORALIE and HARPS.  Following
\citetalias{bouma_wasp4b_2019}, we included the CORALIE measurements
from \citet{wilson_wasp-4b_2008} and \citet{triaud_spin-orbit_2010},
using the homogeneous values calculated by the latter authors.

For HARPS, we included the values reported by
\citet{pont_determining_2011} and \citet{husnoo_observational_2012}.
While \citet{triaud_spin-orbit_2010} also acquired HARPS data over
three nights for Rossiter-McLaughlin observations, these data were
reduced with a non-standard pipeline, and so are systematically offset
from the remaining HARPS data.  In our final fit, we therefore omitted
the three \citet{triaud_spin-orbit_2010} nights.  We did nonetheless
experiment with using the recent \citet{trifonov_public_2020}
re-reductions, which allowed us to include two calibration RVs taken
as part of the \citet{triaud_spin-orbit_2010} program. The decision
regarding whether to include or omit these points did not
noticeably affect our results.

@@

An initial analysis of the new HIRES observations led to our detection
of a linear trend in the residuals after fitting out the orbit of
WASP-4b.  This prompted us to acquire speckle images using Zorro at
Gemini-South \citep[see][and the instrument
web-pages\footnote{\url{www.gemini.edu/sciops/instruments/alopeke-zorro/}}]{scott_nessi_2018}.
Zorro is a dual-channel speckle interferometer employing narrow-band
filters centered at 562$\,$nm and 832$\,$nm.

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29.  Over each observation, we acquired three sets of 1000
60$\,$msec exposures.  If a companion is present, the autocorrelation
functions of these speckle images reveal a characteristic interference
pattern. This pattern in then used to determine the properties of the
detected companion and to produce a reconstructed image.  Using the
reconstructed speckle images,  contrast curves are produced yielding
5-sigma detection limits  (see \citealt{howell_speckle_2011}).  No
companions were detected, and the second night, which had better
seeing (0.6$''$), also produced the more constraining result.  The
832$\,$nm limits were the most useful to us given that faint
companions are redder than the host star.  We therefore opted to use
the 832$\,$nm September 28-29 contrast limits for the remaining
analysis.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.7\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed transit times minus the
  calculated times assuming a constant orbital period.  More opaque
  points correspond to more precise data.  The $\pm1\sigma$
  uncertainties of the quadratic ephemeris 
  are shown in blue.
  The binned TESS point (yellow star) is the weighted average of 18
  TESS transits and is binned for display purposes only.  The models
  were fitted to all of the individual transit times.
  \label{fig:times}
	}
\end{figure*}

We considered two models for the observed transit times.  The first
model assumes a constant orbital period $P$ on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the integer transit number and $t_0$ is a reference
epoch.  The second model assumes that the period changes at a steady
rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
  \label{eq:quadratic_model}
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the observed times.  This helps
to reduce the covariance between $t_0$ and $P$.  A third possible
model that we did not consider for reasons that will become apparent
is a precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

We fitted each model by assuming a Gaussian likelihood and
sampling over the posterior probability distributions.  We sampled the
posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013} in \texttt{emcee}.  The prior for
the quadratic model allowed the period derivative to have any sign.

Figure~\ref{fig:times} shows the observed transit times, minus the
best-fit constant period model.  The best-fitting constant-period
model has 91 degrees of freedom, $\chi^2 = 276$,  and $\chi^2_{\rm
red}=3.0$.  The best-fitting quadratic model has 90 degrees of freedom,
$\chi^2 = 183$, and $\chi^2_{\rm red}=2.0$.  The difference in the
Bayesian information criteria (BIC) between the linear and quadratic
and models is $\Delta {\rm BIC} = 89$, strongly favoring the
quadratic model \citep{kass_bayes_1995}.

From the reduced $\chi^2$ values, we can surmise that neither model
entirely describes the transit data---there
must be some additional source of signal or noise. In
\citetalias{bouma_wasp4b_2019}, we found that the quadratic model for the
(sparser) transit data gave $\chi^2_{\rm red}=1.0$.  The worsened
$\chi^2_{\rm red}$ could reflect underestimated statistical
uncertainties in any of our transit measurements.  It could also
reflect systematic errors in the time-systems in which the transit
measurements were recorded, though we have taken every caution against
this latter possibility.

One approach to analyzing such data would be to inflate the transit
measurement uncertainties, and lower the reduced $\chi^2$.  We do not
think that such an approach is warranted, because it would not change
the result that the quadratic model is strongly preferred.  Instead,
we opt to simply inflate the uncertainties reported for each model by
a factor of $(\chi^2_{\rm red})^{1/2}$, or $\approx$1.73$\times$ for the
linear model, and $\approx$1.41$\times$ for the quadratic.

The resulting best-fit period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within $1\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is $\approx$2.3$\sigma$ larger than the rate
of period decrease we previously reported \citepalias[$- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$;][]{bouma_wasp4b_2019}, presumably because of the
new data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}  The other best-fit transit timing model
parameters are reported in Table~3.

>>>>

\title{WASP-4 is Accelerating Towards the Sun}
% ", Explaining its Decreasing Orbital Period"
% ": A Cautionary Tale in the Search for Tidal Orbital Decay"
% "An Explanation for WASP-4's Decreasing Orbital Period".
% "WASP-4's Decreasing Orbital Period: Solved."

@@

\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be decreasing at
  a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.
  Proposed explanations for the period change include
  tidal orbital decay, apsidal precession, and acceleration of the system
  along the line of sight.
  To investigate further, we performed new radial velocity measurements and
  speckle imaging of WASP-4.
  The radial-velocity data show that the system is
  accelerating towards the Sun at a rate of
  $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
  The associated Doppler effect would cause
  the apparent period to shrink at a rate of
  $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.
  Thus, the observed change in the transit period is mostly
  or entirely produced by the line-of-sight acceleration of the system.
  This acceleration is probably caused by a wide-orbiting companion
  of mass 10-200$\,M_{\rm Jup}$ and orbital distance 10-100$\,$AU,
  based on the magnitude of the radial-velocity trend and the non-detection of any
  companion in the speckle image.  We expect that 1 out of 6 hot Jupiters
  will show apparent period changes of the same order of magnitude as WASP-4,
  based on the hot-Jupiter companion statistics of  
  \citet{knutson_friends_2014}. Continued radial velocity
  monitoring of hot Jupiters is therefore essential to distinguish
  the effects of tidal orbital decay or apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

The orbits of most hot Jupiters are formally unstable to tidal decay
\citep{counselman_outcomes_1973,
hut_stability_1980,rasio_tidal_1996,levrard_falling_2009,matsumura_tidal_2010}.
It is not clear, though, whether the timescale for tidal orbital decay is
shorter or longer than the timescale for main-sequence stellar
evolution.  This answer to this question depends on the uncertain rate at which
friction inside the star damps the tidal oscillations, which is ultimately
what causes the orbit to shrink
(as reviewed by \citealt{Mazeh2008} and
\citealt{ogilvie_tidal_2014}).
Population studies of hot Jupiters --- based on ages, rotation rates,
orbital distances, and Galactic kinematics --- have led to differing
conclusions, with estimated lifetimes ranging from less than a gigayear
to much longer than main-sequence lifetimes
(see, {\it e.g.},
\citealt{jackson_observational_2009}, \citealt{teitler_why_2014},
\citealt{penev_empirical_2018}, \citealt{cameron_hierarchical_2018},
\citealt{hamer_schlaufman_2019}).

An empirical resolution might be possible through long-term timing
of transits and occultations, seeking evidence for changes in the orbital
period.  For instance, long-term transit timing and
radial velocity measurements for WASP-12b have revealed a secular
decrease in the period at a rate of $\approx$30 milliseconds per year, which has
been interpreted as the effect of tidal orbital decay
\citep{maciejewski_departure_2016,patra_2017,yee_orbit_2020}.

This study draws attention to a confounding factor that, while elementary,
does not seem to have
received the attention it deserves.
The point is that observational programs
aimed at identifying orbital decay in hot Jupiters through transit
timing must be accompanied by concurrent long-term radial velocity
monitoring.  The reason is that an apparent change in period
can be produced by the Doppler effect associated with acceleration
of the hot-Jupiter host star along the line of sight, such as the acceleration that
might be produced by a massive, wide-orbiting companion.
Massive outer companions to hot Jupiters are common.
\citet{bryan_statistics_2016} calculated an
occurrence rate of $60.9^{+5.2}_{-5.6}\%$ for outer companions to hot
Jupiters with masses from 1-20$\,$$M_{\rm Jup}$ and semi-major axes
from 5-100$\,$AU.  Therefore, we expect that many hot Jupiters will display
secular trends in orbital period that are unrelated to tidal orbital decay.
This possibility can be checked by performing long-term radial-velocity monitoring
at a level sensitive enough to detect or rule out the relevant amplitude of
acceleration.

The focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that appears to be decreasing by about 10 milliseconds
per year.  The period decrease was identified by \citet{bouma_wasp4b_2019}]{bouma_wasp4b_2019},
who combined data from
the NASA TESS mission \citep{ricker_transiting_2015} and a decade of ground-based
transit observations. Soon thereafter, \citet{southworth_transit_2019} reported an additional
22 transit times and recalculated the period derivative to be $\dot{P} = -9.2 \pm 1.1$
milliseconds per year.  A separate study by \citet{baluev_2019}
reported on additional transit times, and pointed out
that the period decrease was statistically significant only when
analyzing the data with the highest precision.

To determine the origin of the period change, we acquired four
additional radial velocity measurements with the Keck I 10m telescope
and the High Resolution Echelle
Spectrometer (HIRES; \citealt{vogt_hires_1994}).
In doing so, we extended
the time baseline of HIRES measurements from 3 to 9 years.
The previously available HIRES data led to the marginal ($\approx$$2\sigma$)
detection of a radial-velocity trend \citep{knutson_friends_2014}.
Our new measurements reveal a
line-of-sight acceleration of $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m\,s^{-1}\,day^{-1}}$.
Through the Doppler effect\footnote{While the apparent period change caused by a
line-of-sight acceleration has been referred to as the ``R{\o}mer
effect'' \citep{yee_orbit_2020}, a simpler and probably better term is
the Doppler effect. The R{\o}mer effect refers to the {\it delay} in
the reception of a signal due to a change in the time required for
light to traverse the distance between the source and observer. The Doppler
effect refers to the change in the apparent {\it rate} of a process due
to changes in the relative motion of the source and observer, such
as the rate of transits.}, this translates into an expected period decrease
$-5.9$ milliseconds per year, which is comparable to the period decrease
that was measured from transit timing.  We undertook high-resolution
(speckle) imaging to search for evidence of a companion that could
be responsible for the acceleration of WASP-4.

Section~\ref{sec:observations} of this paper presents all of the
available transit data as well as the new radial velocity and
speckle imaging observations.  Section~\ref{sec:analysis} describes
our analysis of the data, and our interpretation
that WASP-4 is being pulled around by a brown dwarf or low-mass star.
Section~\ref{sec:discussion} places this
result within the context of orbital decay searches, and points
out that line-of-sight accelerations will be a relatively
common type of ``false positive.''
Section~\ref{sec:conclusions} offers concluding remarks.

@@

Table~1 lists the transit times we collected for our analysis.  We
included data from the peer-reviewed literature for which {\it (i)} the
reported time was based on the data from a single transit (as opposed
to fitting the data spanning multiple transits and assuming the
period to be constant), {\it (ii)} the
central transit time was allowed to be a free parameter, and {\it (iii)} the time
system was documented clearly, in particular specifying
whether barycentric or heliocentric corrections had been
performed and whether leap seconds had been taken into account (TDB or UTC).

% FIXME : verify with Michael Gillon that this is true. (P. Evans
% done)
Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m telescope at La Silla, Chile,
and the South African Astronomical Observatory 1.0m telescope.
% Five of the twenty-two occurred at simultaneous epochs as transits
% previously reported by \citet{hoyer_tramos_2013}.
Additional timing measurements were also reported recently by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations.
We included twelve of their ``high quality'' transit times from
the TRAPPIST telescope (six transits), El~Sauce (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that correct barycentric and leap-second
corrections had been performed (M.~Gillon, P.~Evans, priv{.} comm{.}).
We omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had been performed.
We did not include in our analysis the four occultation times tabulated by
\citetalias{bouma_wasp4b_2019}, because of the large timing
uncertainties and negligible statistical power.

@@

We acquired four new radial
velocity measurements with Keck/HIRES.  Our observations
were performed using the standard setup and reduction techniques of the
California Planet Survey \citep{howard_cps_2010}.  Previously, the
HIRES data-points spanned 2010 to 2013 \citep{knutson_friends_2014}.
Our new measurements triple the HIRES observing baseline to nine
years.

The complete set of radial velocity observations is given in Table~2.
Along with the 2010--2019 HIRES observations are early
measurements with two different spectrographs.
We included data from CORALIE ({\bf Spell it out on first use)}
from \citet{wilson_wasp-4b_2008} and \citet{triaud_spin-orbit_2010},
adopting the values from the homogeneous analysis of the latter authors.
We included data from HARPS ({\bf spell it out})
from \citet{pont_determining_2011} and \citet{husnoo_observational_2012}.
While \citet{triaud_spin-orbit_2010} also acquired HARPS data over
three nights for Rossiter-McLaughlin observations, these data were
reduced with a non-standard pipeline making them ill-suited
for our study, and we did not include them.\footnote{This problem was
fixed in principle by \citet{trifonov_public_2020} who performed
a homogeneous re-reduction of the entire HARPS data archive.
We found that the decision
regarding whether to include or omit these points did not
noticeably affect our results.}

@@

Once we saw that the new HIRES observations implied a highly significant
trend in the radial velocity,
we sought independent evidence for a wide-orbiting companion by performing
speckle imaging with the Zorro instrument on the 
Gemini South 8m telescope\citep[see][and the instrument
web-pages\footnote{\url{www.gemini.edu/sciops/instruments/alopeke-zorro/}}]{scott_nessi_2018}.
Zorro is a dual-channel speckle interferometer employing narrow-band
filters centered at 562$\,$nm and 832$\,$nm.

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29.  On each night, we acquired three sets of $1000\times 60$\,$msec exposures.
If a companion is present, the autocorrelation
functions of these speckle images would reveal a characteristic interference
pattern. This pattern in then used to determine the properties of the
detected companion and produce a reconstructed image.  Using the
reconstructed speckle images, contrast curves are produced to determine the
5-$\sigma$ detection limits (see \citealt{howell_speckle_2011}).  No
companions were detected. The data from the second night, which had better
seeing (0.6$''$), led to the most constraining limit.  The
832$\,$nm limits were the most useful, given that any faint
companion would likely be redder than the host star.  Therefore,
we opted to use
the 832$\,$nm September 28-29 contrast limits in the analysis described below.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.7\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed transit times minus the
  calculated times, assuming a constant orbital period.  More opaque
  points correspond to more precise data.  The $\pm1\sigma$
  uncertainties of the quadratic ephemeris 
  are shown in blue.
  The yellow star represents the weighted average of 18 data points
  obtained with TESS. The TESS data were averaged here for display purposes only;
  our analysis used the 18 individual transit times.
  \label{fig:times}
	}
\end{figure*}

We fitted two simple timing models to the transit-timing data. The first
model assumes the period $P$ to be constant:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the integer specifying the number within the sequence of orbits
spanned by the data,
and $t_0$ is the transit time for the event designated $E=0$.
The second model assumes that the period changes at a steady
rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
  \label{eq:quadratic_model}
\end{align}
The free parameters are the reference time $t_0$, the period at the
reference epoch $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the times of observed transits.
This choice leads to a small covariance between the uncertainties in $t_0$ and $P$.
%A third possible
%model that we did not consider for reasons that will become apparent
%is a precessing, eccentric orbit \citep[{\it
%e.g.},][]{gimenez_revision_1995,patra_2017}.

We fitted each model by assuming a Gaussian likelihood and
sampling over the posterior probability distributions.  We sampled the
posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013} in \texttt{emcee}.  The prior for
the quadratic model allowed the period derivative to be either positive
or negative.

Figure~\ref{fig:times} shows the observed transit times, minus the
best-fitting constant-period model.  The best-fitting constant-period
model has 91 degrees of freedom, $\chi^2 = 276$, and $\chi^2_{\rm
red}=3.0$.  The best-fitting quadratic model has 90 degrees of freedom,
$\chi^2 = 183$, and $\chi^2_{\rm red}=2.0$.  The difference in the
Bayesian information criteria (BIC) between the linear and quadratic
and models is $\Delta {\rm BIC} = 89$, strongly favoring the
quadratic model \citep{kass_bayes_1995}.

From the high values of $\chi^2_{\rm red}$, we can surmise that neither model
provides a satisfactory fit to the transit data; there
must be some additional signal or noise.
The earlier study by \citetalias{bouma_wasp4b_2019} found that the quadratic model for the
(sparser) transit data gave $\chi^2_{\rm red}=1.0$.  The worsening 
of $\chi^2_{\rm red}$ probably reflects underestimated measurement
uncertainties in at least some of the newly reported transit times,
due to time correlated noise in the photometry. It might also be due
to errors or misunderstandings of the time systems in which the
data were recorded, despite our best efforts to interpret the literature.
In what follows, we opted to enlarge the uncertainties in the best-fitting parameters
of each model by a factor of $(\chi^2_{\rm red})^{1/2}$.  This was a factor of $\approx$1.73$\times$ for the
linear model, and $\approx$1.41$\times$ for the quadratic model.

The resulting period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within 1-$\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is 2.3-$\sigma$ larger than the rate
of period decrease reported by \citepalias[][; $- 12.6 \pm 1.2~{\rm ms}\,{\rm yr}^{-1}$]{bouma_wasp4b_2019},
presumably because of the new data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}. The other best-fitting model
parameters are reported in Table~3.
